---
title: "Pipeline QC metrics"
author: "Mike Connelly"
date: "07/29/2025"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```
```{r packages, message=FALSE}
library("tidyverse")
#
source("./R/setup_colors_functions.R")
```

```{r sample_data_import}
samples <- read_csv("./data/Towson_Hydractiniidae_all_metadata_march2026.csv")
```

## Import/inspect genome skimming pipeline QC metrics
  Note: Run MultiQC on R1 reads only to create reports for all 358 analysis samples (125 + 233).
        
### Raw reads
```{r import_multiqc_data}
# raw reads
multiqc_fastqc <- read_tsv("./outputs/QCs/fastqcs/multiqc_data/multiqc_fastqc.txt") %>%
  mutate("Sample_ID" = gsub("_R[12]", "", Sample)) %>%
  left_join(samples, by = "Sample_ID") %>% 
  mutate("Site code" = ordered(`Site code`, levels = names(site_cols)))
```
```{r inspect_read_depth}
# quick check of sample depth
pdf("./outputs/figures/total_read_distribution.pdf", width = 6.5, height = 4.76)
multiqc_fastqc %>%
  mutate(Sample = fct_reorder(Sample, desc(`Total Sequences`))) %>%
  ggplot(aes(Sample, `Total Sequences`, color = `Site code`)) +
  scale_color_manual(values = site_cols) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 1000000, color = "red") +
  ylim(c(0, 1e8)) +
  theme(axis.text.x = element_blank()) +
  ggtitle("Total number of Raw Reads per sample")
dev.off()
```

```{r read_stats}
# Produce total read summary statistics for manuscript
sum(multiqc_fastqc$`Total Sequences`)
# 17,684,433,041
mean(multiqc_fastqc$`Total Sequences`)
# 49,397,858
sd(multiqc_fastqc$`Total Sequences`)
# 14,660,208
```
```{r}
multiqc_fastqc %>% group_by(Round) %>% summarise("total" = sum(`Total Sequences`),
                                                 "mean" = mean(`Total Sequences`))
multiqc_fastqc %>% group_by(Batch) %>% summarise("total" = sum(`Total Sequences`),
                                                 "mean" = mean(`Total Sequences`))
```

```{r}
pdf("./outputs/figures/read_site_distribution.pdf", width = 8.5, height = 4.76)
multiqc_fastqc %>% 
  ggplot(aes(`Site code`, `Total Sequences`, color = `Site code`)) +
  geom_boxplot() +
  geom_point() +
  scale_color_manual(values = site_cols) +
  geom_hline(yintercept = 1e7, lty = 2) +
  ylim(c(0, 1.2e8)) +
  theme(legend.position = "none") +
  ggtitle("Raw Read Pairs per Site")
dev.off()
```

```{r}
multiqc_fastqc %>% ggplot(aes(total_deduplicated_percentage)) + geom_histogram(bins = 50)
multiqc_fastqc %>% ggplot(aes(`%GC`, fill = `Site code`)) + geom_histogram(binwidth = 0.5)
```


### Trimmed reads - Trimmomatic vs FASTP

### Aligned reads
```{r import_multiqc_data}
# raw reads
multiqc_flagstat <- read_tsv("./outputs/QCs/flagstats/multiqc_data/multiqc_samtools_flagstat.txt") %>%
  mutate("Sample_ID" = gsub("_flagstat", "", Sample)) %>%
  left_join(samples, by = "Sample_ID") %>% 
  mutate("Site code" = ordered(`Site code`, levels = names(site_cols)))
```
```{r}
multiqc_flagstat %>%
  mutate(Sample = fct_reorder(Sample, desc(`flagstat_total`))) %>%
  ggplot(aes(Sample, `properly paired_passed_pct`, color = `Site code`)) +
  geom_point(size = 1.5) +
  scale_color_manual(values = site_cols) +
  geom_hline(yintercept = 35, color = "red") +
  ylim(c(0, 100)) +
  theme(axis.text.x = element_blank()) +
  ggtitle("Mapping rate")
```
```{r alignment_stats}
# Produce alignment summary statistics for manuscript
mean(multiqc_flagstat$`mapped_passed_pct`)
# 83.86264
sd(multiqc_flagstat$`mapped_passed_pct`)
# 18.55942
range(multiqc_flagstat$`mapped_passed_pct`)
# 14.40 98.97
```
```{r}
pdf("./outputs/figures/mapping_site_distribution.pdf", width = 6.5, height = 4.76)
multiqc_flagstat %>% 
  ggplot(aes(`Site code`, `mapped_passed_pct`, color = `Site code`)) +
  geom_boxplot() +
  geom_point() +
  scale_color_manual(values = site_cols) +
  ylim(c(0, 100)) +
  ggtitle("Read mapping percentages per site")
dev.off()
```
Explore correlations between read parameters (GC content, duplicate reads, warning flags) and mapping rates, also explore reasons for outlier samples in Pcangsd --> cutoffs for alignment rates, SNP completeness, etc?
