---
title: "Sharkmer barcodes"
author: "Mike Connelly"
date: "2025-09-24"
output: html_document
---
## Setup and packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
options(stringsAsFactors = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```
```{r packages, message=FALSE}
library("tidyverse")
library("ape")
library("pegas")
#
source("./R/setup_colors_functions.R")
#
`%notin%` <- Negate(`%in%`)
```

## Import sample metadata
```{r sample_data_import}
samples <- read_csv("./data/Towson_Hydractiniidae_all_metadata_march2026.csv")
```
```{r}
blastn_cols <- c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore", "sscinames", "stitle") #
```

## Analyze barcode sequences obtained using sharkmer
Mitochondrial: 16S, CO1 
Nuclear: 18S, 28S(-v2), ITS(-v2), EF1A

 - Inspect sharkmer results
 - Best BLASTN hits
 - Multiple sequence alignment
 - Haplotype network (16S, CO1)
 - Phylogenetic tree

Which samples do I have which barcodes for? Presence/absence table

## Miglietta 16S
  miglietta_16S
    forward: ACGGAATGAACTCAAATCATGT
    reverse: TCGACTGTTTACCAAAAACATA
    max-length: 1000
    min-coverage: 2
    mismatches: 2
    trim: 15
    citation: Miglietta et al. 2009
    notes: Product is about 700bp long
    arguments: --pcr "forward=ACGGAATGAACTCAAATCATGT,reverse=TCGACTGTTTACCAAAAACATA,max-length=1000,name=miglietta_16S"

### Sharkmer results
```{r}
samples_miglietta_16S_sharkmer <- read_delim("./outputs/sharkmer/all_processed/miglietta_16S/sharkmer_all_miglietta_16S_fasta_summary.txt") %>%
  mutate(`Sample_ID` = gsub("_mig.*$", "", gsub(">", "", seq))) %>%
  left_join(samples)

samples_miglietta_16S_sharkmer %>% group_by(Sample_ID) %>% count() %>% View()
```

### Import BLAST hits
```{r}
blast_hits_miglietta_16S <- read_tsv("./outputs/sharkmer/all_processed/miglietta_16S/sharkmer_all_miglietta_16S_600_top_blastn_hit.txt", col_names = blastn_cols)

best_blast_hits_miglietta_16S <- blast_hits_miglietta_16S %>% group_by(qseqid) %>% slice_max(bitscore)

best_blast_hits_miglietta_16S %>% group_by(sseqid, stitle) %>% count()

# 347/358 full-length sequences (96.9%), hits to H. symbiolongicarpus (312), H. polyclina (33), P. americana (1), and Eucheilota maculata (1) mt genomes
# missing: AST_12, CCB_10, CCB_11, LHP_5, WHS_3B, BSH_9, OSB_5, OSB_10A, OSB_15, ECJ_1
```
```{r}
sample_species_IDs <- best_blast_hits_miglietta_16S %>% mutate("Sample_ID" = sapply(str_split(qseqid, "_"), function(x) paste(x[1], x[2], sep = "_")),
                                         "Species_ID" = ifelse(grepl("symbiolongicarpus", stitle), "Hydractinia symbiolongicarpus",
                                                               ifelse(grepl("polyclina", stitle), "Hydractinia polyclina",
                                                                      ifelse(grepl("Podocoryna", stitle), "Podocoryna americana",
                                                                             "Eucheilota maculata")))) %>% 
  dplyr::select(Sample_ID, Species_ID)
```
```{r}
sample_species_IDs %>% filter(Species_ID == "Hydractinia polyclina") %>% View() #%>% select(Sample_ID) %>% write_lines("./data/hpoly_samples.txt")
```

### Import sequence alignment 
```{r import_alignment}
data <- read.dna("./outputs/sharkmer/all_processed/miglietta_16S/sharkmer_midatlantic_miglietta_16S_600_aligned_clipkit.fasta", format = "fasta")
checkAlignment(data, what = 1:4)
# 120 sequences, 631 sites
# attributes(data)
# dimnames(data)[[1]]
```

### Haplotype network 
```{r haplotypes}
dataHaplo <- haplotype(data)
dataHaplo
```

```{r extract haplotypes and build network}
dataHaplo <- sort(dataHaplo, what = "labels")
# attributes(dataHaplo)
dataNet <- haploNet(dataHaplo)
# dataNet <- mst(dataHaplo)
# dataNet <- mjn(dataHaplo)
```
```{r basic haplotype net}
plot(dataNet,
     show.mutation = 3)
```
```{r}
plot(dataNet,
     size = attr(dataNet, "freq"),
     show.mutation = 3,
     labels = F,
     legend = c(-100,100))
```
```{r extract sequence labels}
seq_labels <- dimnames(data)[[1]]
# seq_labels
seq_ids <- gsub("_mig.*$", "", seq_labels)
# seq_ids
```
```{r countHap}
countHap <- function(hap = h, dna = x){
  with(
    stack(setNames(attr(hap, "index"), rownames(hap))),
    table(haplotype = ind, seq = attr(dna, "dimnames")[[1]][values])
  )
}
```
```{r genotype haplotype matches}
df <- as.data.frame(countHap(dataHaplo, data))
good <- df %>% filter(Freq == 1)
```
```{r select_good_haplotypes}
good$`Sample_ID` <- gsub("_mig.*", "", good$seq)
#
sample_haplotypes <- good %>% dplyr::select(`Sample_ID`, `haplotype`) %>% left_join(samples)
```

### Summarize haplotypes
```{r}
sample_haplotypes %>% group_by(haplotype, `Site code`) %>% count() %>% pivot_wider(names_from = `Site code`, values_from = n)
```

### Phylogenetic tree

Include only unique haplotypes and reference sequences

## Cnidaria 16S
  cnidaria_16S
    forward: GRCTGTTTACCAAAAACATA
    reverse: AATTCAACATMGAGG
    min-length: 500
    max-length: 700
    min-coverage: 2
    mismatches: 2
    trim: 15
    citation: Modified from Cunningham and Buss 1993 https://doi.org/10.1016/0305-1978(93)90009-G 
    notes: Amplifies portions of domains IV and V. Product is about 580bp
    arguments: --pcr "forward=GRCTGTTTACCAAAAACATA,reverse=AATTCAACATMGAGG,name=cnidaria_16S,max-length=700,min-length=500,min-coverage=2,mismatches=2,trim=15"

### Sharkmer results
```{r}
samples_cnidaria_16S_sharkmer <- read_delim("./outputs/sharkmer/all_processed/cnidaria_16S/sharkmer_all_cnidaria_16S_fasta_summary.txt") %>%
  mutate(`Sample_ID` = gsub("_cnid.*$", "", gsub(">", "", seq))) %>%
  left_join(samples)

samples_cnidaria_16S_sharkmer %>% group_by(Sample_ID) %>% count()
```

### Import BLAST hits
```{r}
blast_hits_cnidaria_16S <- read_tsv("./outputs/sharkmer/all_processed/cnidaria_16S/sharkmer_all_cnidaria_16S_500_top_blastn_hit.txt", col_names = blastn_cols)

best_blast_hits_cnidaria_16S <- blast_hits_cnidaria_16S %>% group_by(qseqid) %>% slice_max(bitscore, with_ties = F)

best_blast_hits_cnidaria_16S %>% group_by(sseqid, stitle) %>% count()

# 352/358 full-length sequences (96.9%), hits to H. symbiolongicarpus (311), H. polyclina (33), P. americana (1), and Eucheilota maculata (1) mt genomes
```


## Cnidaria CO1
  cnidaria_CO1
    forward: TCATAARGATATHGG
    reverse: RTGNCCAAAAAACCA
    min-length: 600
    max-length: 800
    min-coverage: 2
    mismatches: 2
    trim: 15
    citation: Modified from Folmer et al. 1994
    notes: Product is about 700bp long
    arguments: --pcr "forward=TCATAARGATATHGG,reverse=RTGNCCAAAAAACCA,name=cnidaria_CO1,max-length=800,min-length=600,min-coverage=2,mismatches=2,trim=15"

### Sharkmer results
```{r}
samples_cnidaria_CO1_sharkmer <- read_delim("./outputs/sharkmer/all_processed/cnidaria_CO1/sharkmer_all_cnidaria_CO1_fasta_summary.txt") %>%
  mutate(`Sample_ID` = gsub("_cni.*$", "", gsub(">", "", seq))) %>%
  left_join(samples)

samples_cnidaria_CO1_sharkmer %>% group_by(Sample_ID) %>% count() %>% View()
```

### Import BLAST hits
```{r}
blast_hits_cnidaria_CO1 <- read_tsv("./outputs/sharkmer/all_processed/cnidaria_CO1/sharkmer_all_cnidaria_CO1_600_top_blastn_hit.txt", col_names = blastn_cols)

best_blast_hits_cnidaria_CO1 <- blast_hits_cnidaria_CO1 %>% group_by(qseqid) %>% slice_max(bitscore)

best_blast_hits_cnidaria_CO1 %>% group_by(sseqid, stitle) %>% count()

```

### Import sequence alignment 
```{r import_alignment}
data <- read.dna("./outputs/sharkmer/all_processed/sharkmer_midatlantic_cnidaria_CO1_600_aligned_clipkit.fasta", format = "fasta")
checkAlignment(data, what = 1:4)
# 120 sequences, 631 sites, gaps
# attributes(data)
# dimnames(data)[[1]]
```


## Cnidaria 18S
  cnidaria_18S
    forward: AACCTGGTTGATCCTGCCAGT
    reverse: TGATCCTTCTGCAGGTTCACCTAC
    min-length: 1600
    max-length: 2000
    min-coverage: 2
    mismatches: 2
    trim: 15
    citation: Medlin et al. 1988 https://doi.org/10.1016/0378-1119(88)90066-2
    notes: Product is about 1790bp
    arguments: --pcr "forward=AACCTGGTTGATCCTGCCAGT,reverse=TGATCCTTCTGCAGGTTCACCTAC,name=cnidaria_18S,max-length=2000,min-length=1600,min-coverage=2,mismatches=2,trim=15"
    
### Sharkmer results
```{r}
samples_cnidaria_18S_sharkmer <- read_delim("./outputs/sharkmer/all_processed/cnidaria_18S/sharkmer_all_cnidaria_18S_fasta_summary.txt") %>%
  mutate(`Sample_ID` = gsub("_cni.*$", "", gsub(">", "", seq))) %>%
  left_join(samples)

samples_cnidaria_18S_sharkmer %>% group_by(Sample_ID) %>% count() %>% View()
```

### Import BLAST hits
```{r}
blast_hits_cnidaria_18S <- read_tsv("./outputs/sharkmer/all_processed/cnidaria_18S/sharkmer_all_cnidaria_18S_1500_top_blastn_hit.txt", col_names = blastn_cols)
best_blast_hits_cnidaria_18S <- blast_hits_cnidaria_18S %>% group_by(qseqid) %>% slice_max(bitscore, with_ties = F)

best_blast_hits_cnidaria_18S %>% group_by(sseqid, stitle) %>% count()

```

## Cnidaria 28S
  cnidaria_28S
    forward: CCYYAGTAACGGCGAGT
    reverse: SWACAGATGGTAGCTTCG
    min-length: 2900
    max-length: 3500
    min-coverage: 2
    mismatches: 2
    trim: 15
    citation: Evans et al. 2008. https://doi.org/10.1186/1471-2148-8-139
    notes: Forward primer is F97, reverse is R3238. Product is about 3230bp.
    arguments: --pcr "forward=CCYYAGTAACGGCGAGT,reverse=SWACAGATGGTAGCTTCG,name=cnidaria_28S,max-length=3500,min-length=2900,min-coverage=2,mismatches=2,trim=15"
    
### Sharkmer results
```{r}
samples_cnidaria_28S_sharkmer <- read_delim("./outputs/sharkmer/all_processed/cnidaria_28S/sharkmer_all_cnidaria_28S_fasta_summary.txt") %>%
  mutate(`Sample_ID` = gsub("_cni.*$", "", gsub(">", "", seq))) %>%
  left_join(samples)

samples_cnidaria_28S_sharkmer %>% group_by(Sample_ID) %>% count() %>% View()
```
    
### Import BLAST hits
```{r}
blast_hits_cnidaria_28S <- read_tsv("./outputs/sharkmer/all_processed/cnidaria_28S/sharkmer_all_cnidaria_28S_1000_top_blastn_hit.txt", col_names = blastn_cols)
best_blast_hits_cnidaria_28S <- blast_hits_cnidaria_28S %>% group_by(qseqid) %>% slice_max(bitscore, with_ties = F)

best_blast_hits_cnidaria_28S %>% group_by(sseqid, stitle) %>% count()

```

## Cnidaria 28S-v2
  cnidaria_28S-v2
    forward: CGTGAAACCGYTRRAAGGG
    reverse: TTGGTCCGTGTTTCAAGACG
    min-length: 300
    max-length: 700
    min-coverage: 2
    mismatches: 2
    trim: 15
    citation: McCartin et al 2024. https://doi.org/10.7717/peerj.18607
    notes: Anth-28S-eDNA; Product is about 540bp
    arguments: --pcr "forward=CGTGAAACCGYTRRAAGGG,reverse=TTGGTCCGTGTTTCAAGACG,name=cnidaria_28S-v2,max-length=700,min-length=300,min-coverage=2,mismatches=2,trim=15"
    
### Sharkmer results
```{r}
samples_cnidaria_28Sv2_sharkmer <- read_delim("./outputs/sharkmer/all_processed/cnidaria_28S-v2/sharkmer_all_cnidaria_28S-v2_fasta_summary.txt") %>%
  mutate(`Sample_ID` = gsub("_cni.*$", "", gsub(">", "", seq))) %>%
  left_join(samples)

samples_cnidaria_28Sv2_sharkmer %>% group_by(Sample_ID) %>% count() %>% View()
```
    
### Import BLAST hits
```{r}
blast_hits_cnidaria_28Sv2 <- read_tsv("./outputs/sharkmer/all_processed/cnidaria_28S-v2/sharkmer_all_cnidaria_28S-v2_400_top_blastn_hit.txt", col_names = blastn_cols)
best_blast_hits_cnidaria_28Sv2 <- blast_hits_cnidaria_28Sv2 %>% group_by(qseqid) %>% slice_max(bitscore, with_ties = F)

best_blast_hits_cnidaria_28Sv2 %>% group_by(sseqid, stitle) %>% count()

```

## Cnidaria ITS
  cnidaria_ITS
    forward: TACACACCGCCCGTCGCTACTA
    reverse: ACTCGCCGTTACTRRGG
    min-length: 600
    max-length: 1000
    min-coverage: 2
    mismatches: 2
    trim: 15
    citation: Ji et al 2003. https://doi.org/10.1046/j.1471-8286.2003.00519.x
    notes: Forward primer is CAS18SF1, position 1843, from Ji et al 2003. The reverse primer is the reverse complement of the 28S forward primer. Product is about 770-880bp.
    arguments: --pcr "forward=TACACACCGCCCGTCGCTACTA,reverse=ACTCGCCGTTACTRRGG,name=cnidaria_ITS,max-length=1000,min-length=600,min-coverage=2,mismatches=2,trim=15"
    
### Sharkmer results
```{r}
samples_cnidaria_ITS_sharkmer <- read_delim("./outputs/sharkmer/all_processed/cnidaria_ITS/sharkmer_all_cnidaria_ITS_fasta_summary.txt") %>%
  mutate(`Sample_ID` = gsub("_cni.*$", "", gsub(">", "", seq))) %>%
  left_join(samples)

samples_cnidaria_ITS_sharkmer %>% group_by(Sample_ID) %>% count() %>% View() #%>% ungroup %>%  count(n) %>%  View()
```

### Import BLAST hits
```{r}
blast_hits_cnidaria_ITS <- read_tsv("./outputs/sharkmer/all_processed/cnidaria_ITS/sharkmer_all_cnidaria_ITS_500_top_blastn_hit.txt", col_names = blastn_cols)
best_blast_hits_cnidaria_ITS <- blast_hits_cnidaria_ITS %>% group_by(qseqid) %>% slice_max(bitscore, with_ties = F)

best_blast_hits_cnidaria_ITS %>% group_by(sseqid, stitle) %>% count()

```
## Cnidaria ITS
  cnidaria_ITS-v2
    forward: GTAGGTGAACCTGCAGAAGGATCA
    reverse: ACTCGCCGTTACTRRGG
    min-length: 600
    max-length: 1000
    min-coverage: 2
    mismatches: 2
    trim: 15
    citation: Ji et al 2003. https://doi.org/10.1046/j.1471-8286.2003.00519.x
    notes: Forward primer is reverse complement of 18S reverse primer. The reverse primer is the reverse complement of the 28S forward primer. Product is about 620-730bp.
    arguments: --pcr "forward=GTAGGTGAACCTGCAGAAGGATCA,reverse=ACTCGCCGTTACTRRGG,name=cnidaria_ITS-v2,max-length=1000,min-length=600,min-coverage=2,mismatches=2,trim=15"
    
### Sharkmer results
```{r}
samples_cnidaria_ITSv2_sharkmer <- read_delim("./outputs/sharkmer/all_processed/cnidaria_ITS-v2/sharkmer_all_cnidaria_ITS-v2_fasta_summary.txt") %>%
  mutate(`Sample_ID` = gsub("_cni.*$", "", gsub(">", "", seq))) %>%
  left_join(samples)

samples_cnidaria_ITSv2_sharkmer %>% group_by(Sample_ID) %>% count() %>% View() #%>% ungroup %>%  count(n) %>%  View()
```

### Import BLAST hits
```{r}
blast_hits_cnidaria_ITSv2 <- read_tsv("./outputs/sharkmer/all_processed/cnidaria_ITS-v2/sharkmer_all_cnidaria_ITS-v2_500_top_blastn_hit.txt", col_names = blastn_cols)
best_blast_hits_cnidaria_ITSv2 <- blast_hits_cnidaria_ITS %>% group_by(qseqid) %>% slice_max(bitscore)

best_blast_hits_cnidaria_ITSv2 %>% group_by(sseqid, stitle) %>% count()

```


## Cnidaria EF1A
  cnidaria_EF1A
    forward: AMGWGGHATGGTDGCTGGTG
    reverse: YTTRATNAYDCCAACAGCWAC
    min-length: 200
    max-length: 3000
    min-coverage: 2
    mismatches: 2
    trim: 15
    citation: 
    notes: 
    arguments: --pcr "forward=AMGWGGHATGGTDGCTGGTG,reverse=YTTRATNAYDCCAACAGCWAC,name=cnidaria_EF1A,max-length=3000,min-length=200,min-coverage=2,mismatches=2,trim=15"
    
### Sharkmer results
```{r}
samples_cnidaria_EF1A_sharkmer <- read_delim("./outputs/sharkmer/all_processed/cnidaria_EF1A/sharkmer_all_cnidaria_EF1A_fasta_summary.txt") %>%
  mutate(`Sample_ID` = gsub("_cni.*$", "", gsub(">", "", seq))) %>%
  left_join(samples)

samples_cnidaria_EF1A_sharkmer %>% group_by(Sample_ID) %>% count() %>% ungroup %>%  count(n) %>%  View()
```

### Import BLAST hits
```{r}
blast_hits_cnidaria_EF1A <- read_tsv("./outputs/sharkmer/all_processed/sharkmer_all_cnidaria_EF1A_200_top_blastn_hit.txt", col_names = blastn_cols)
best_blast_hits_cnidaria_EF1A <- blast_hits_cnidaria_EF1A %>% group_by(qseqid) %>% slice_max(bitscore, with_ties = F)

best_blast_hits_cnidaria_EF1A %>% group_by(sseqid, stitle) %>% count()

```

